---
title: "Liotta_Dario_Rlab06"
output: html_document
date: "2024-06-07"
---

```{r}
library(rstan)
library(ggplot2)
library(bayesplot)
```

# Exercise 1

```{r}
# Data
X1 <- c(109, 65, 22, 3, 1, 0)
X2 <- c(144, 91, 32, 11, 2, 0)
```

A uniform prior and a Poisson likelihood gives a Gamma posterior $\sim t^{k-1}e^{-\mu t}$ where

$$\begin{cases}k=\sum\limits_{i=1}^nx_i+1\\ \mu=n\end{cases}$$

with mean and variance given by

$$E(\lambda)=\frac{k}{\mu} \hspace{1cm} var(\lambda)=\frac{k}{\mu^2}$$

```{r}
lambdas <- seq(0, 1, by = 10^(-4))

# Parameters for first set of observations
k1 <- 0
for(i in 1:length(X1)) {
  k1 <- k1 + (i-1) * X1[i]
}
mu1 <- sum(X1)
values1  <- dgamma(lambdas, shape = k1, rate = mu1)

# Parameters for second set of observations
k2 <- 0
for(i in 1:length(X2)) {
  k2 <- k2 + (i-1) * X2[i]
}
mu2 <- sum(X2)
values2  <- dgamma(lambdas, shape = k2, rate = mu2)

# Plot
plot(lambdas, values1,
     xlab = expression(lambda),
     ylab = "Probability density",
     main = "Posterior distribution with uniform prior",
     type = "l", lwd = 2, col = "blue",
     ylim = c(0, max(values2))
     )
lines(lambdas, values2, type = "l", lwd = 2, col = "red")
legend("topright", legend = c("First set", "Second set"), col = c("blue", "red"), lwd = 2)
```

```{r}
# Printing results
cat(sprintf("About the first set of observations:\n"))
cat(sprintf("Mean: %.4f\n", k1/mu1))
cat(sprintf("Variance: %.4f\n", k1/(mu1^2)))
cat(sprintf("Median: %.4f\n", qgamma(0.5, shape = k1, rate = mu1)))
cat(sprintf("95%% credibility interval: [%.4f, %.4f]\n", qgamma(0.025, shape = k1, rate = mu1), qgamma(0.975, shape = k1, rate = mu1)))

cat(sprintf("\n\nAbout the second set of observations:\n"))
cat(sprintf("Mean: %.4f\n", k2/mu2))
cat(sprintf("Variance: %.4f\n", k2/(mu2^2)))
cat(sprintf("Median: %.4f\n", qgamma(0.5, shape = k2, rate = mu2)))
cat(sprintf("95%% credibility interval: [%.4f, %.4f]\n", qgamma(0.025, shape = k2, rate = mu2), qgamma(0.975, shape = k2, rate = mu2)))
```

Using a Jeffrey prior changes only the shape parameter, which becomes $k=\sum\limits_{i=1}^nx_i+\frac{1}{2}$, so we can just pass from $k$ to $k-\frac{1}{2}$ and do the same calculations.

```{r}
# Correcting parameters
k1 <- k1 - 0.5
k2 <- k2 - 0.5

values1  <- dgamma(lambdas, shape = k1, rate = mu1)
values2  <- dgamma(lambdas, shape = k2, rate = mu2)

# Plot
plot(lambdas, values1,
     xlab = expression(lambda),
     ylab = "Probability density",
     main = "Posterior distribution with Jeffrey prior",
     type = "l", lwd = 2, col = "blue",
     ylim = c(0, max(values2))
    )
lines(lambdas, values2, type = "l", lwd = 2, col = "red")
legend("topright", legend = c("First set", "Second set"), col = c("blue", "red"), lwd = 2)
```

```{r}
# Printing results
cat(sprintf("About the first set of observations:\n"))
cat(sprintf("Mean: %.4f\n", k1/mu1))
cat(sprintf("Variance: %.4f\n", k1/(mu1^2)))
cat(sprintf("Median: %.4f\n", qgamma(0.5, shape = k1, rate = mu1)))
cat(sprintf("95%% credibility interval: [%.4f, %.4f]\n", qgamma(0.025, shape = k1, rate = mu1), qgamma(0.975, shape = k1, rate = mu1)))

cat(sprintf("\n\nAbout the second set of observations:\n"))
cat(sprintf("Mean: %.4f\n", k2/mu2))
cat(sprintf("Variance: %.4f\n", k2/(mu2^2)))
cat(sprintf("Median: %.4f\n", qgamma(0.5, shape = k2, rate = mu2)))
cat(sprintf("95%% credibility interval: [%.4f, %.4f]\n", qgamma(0.025, shape = k2, rate = mu2), qgamma(0.975, shape = k2, rate = mu2)))
```

# Exercise 2

```{r}
# Data
y1    <- c(rep(0, 109), rep(1, 65), rep(2, 22), rep(3, 3), rep(4, 1))
y2    <- c(rep(0, 144), rep(1, 91), rep(2, 32), rep(3, 11), rep(4, 2))
data1 <- list(N = length(y1), y = y1)
data2 <- list(N = length(y2), y = y2)

# Model strings for stan
model_string_uniform = 
  " 
    data {
      int <lower=0> N; 
      int y[N]; 
    } 
    
    parameters { 
      real <lower=0> Lambda; 
    } 
    
    model { 
      y ~ poisson (Lambda); 
    }
  "

model_string_jeffrey = 
  " 
    data {
      int <lower=0> N; 
      int y[N]; 
    } 
    
    parameters { 
      real <lower=0> Lambda; 
    } 
    
    model { 
      target += -0.5*log(Lambda);
      y ~ poisson (Lambda); 
    }
  "
```

```{r}
# Function for the posterior analysis
posterior_analysis <- function(model_string, data_list, bin_width = 0.01) {
  stan_m  <- stan_model(model_code = model_string)
  stanFit <- sampling(object = stan_m,
                      data = data_list, 
                      chains = 2, 
                      iter = 10000, 
                      warmup = 1000, 
                      thin = 1)
  
  # Plot of histogram
  hist_plot <- mcmc_hist(stanFit, pars = c("Lambda"), binwidth = bin_width) +
               labs(title = "Posterior histogram of λ",
                    x = expression(lambda),
                    y = "Frequency") +
               theme_minimal() +
               theme(plot.title = element_text(hjust = 0.5))
  
  # Print of results
  summary_df <- as.data.frame(summary(stanFit)$summary)
  par_values <- summary_df["Lambda", ]
  
  cat(sprintf("\n\n\n"))
  cat(sprintf("Mean: %.4f\n", par_values["mean"]))
  cat(sprintf("Variance: %.4f\n", par_values["sd"]^2))
  cat(sprintf("Median: %.4f\n", par_values["50%"]))
  cat(sprintf("95%% credibility interval: [%.4f, %.4f]\n", par_values["2.5%"], par_values["97.5%"]))
  
  print(hist_plot)
}
```

Results for the first set of observations (considering a uniform prior).

```{r}
posterior_analysis(model_string_uniform, data1)
```

Results for the second set of observations (considering a uniform prior).

```{r}
posterior_analysis(model_string_uniform, data2)
```

Results for the first set of observations (considering a Jeffrey prior).

```{r}
posterior_analysis(model_string_jeffrey, data1)
```

Results for the second set of observations (considering a Jeffrey prior).

```{r}
posterior_analysis(model_string_jeffrey, data2)
```

# Exercise 3

***(a) find the frequentist estimator for p***

Given $y$ samples with high bacter level on a total of $n$, the frequentist estimator will be given by $\frac{y}{n}$.

```{r}
y <- 11
n <- 116

cat(sprintf("Frequentist estimator for p: %.3f", y/n))
```

***(b) using a Beta(1, 10) prior for p, calculate and posterior distribution P (p y)***

Given a beta prior $P(p|\alpha,\beta)$ and a binomial likelihood $P(y|n,p)$, the posterior will be a beta of the form $P(p|\alpha+y,\beta+n-y)$, so we find a posterior:

$$Beta(p|1+11,10+116-11)=Beta(p|12,115)$$

***(c) find the bayesian estimator for p, the posterior mean and variance, and a 95% credible interval***

The bayesian estimator will be given by the mean of the posterior distribution, which, for a beta distribution, is

$$E(p)=\frac{\alpha}{\alpha+\beta}$$

```{r}
alpha <- 12
beta  <- 115

cat(sprintf("Bayesian estimator for p: %.3f", alpha/(alpha+beta)))
```

The variance is

$$var(p)=\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$$

```{r}
var_p <- (alpha*beta) / ((alpha+beta)*(alpha+beta)*(alpha+beta+1))

cat(sprintf("Variance: %.5f", var_p))
```

Lastly, let's find the $95%$ credibility interval.

```{r}
cat(sprintf("95%% credibility interval: [%.3f, %.3f]", qbeta(0.025, alpha, beta), qbeta(0.975, alpha, beta)))
```

***(d) test the hypothesis***

$$H_0:p=0.1 \ \ \text{versus} \ \ H_1:p\neq0.1$$

***at 5% level of significance with both the frequentist and bayesian approach***

```{r}
binom.test(x = y, n = n, p = 0.1)
```

Since $0.1$ also falls into the $95%$ credibility interval for the posterior distribution, we accept the hypothesis with both the frequentist and the bayesian approach.

***(e) find the frequentist estimator for p***

```{r}
y <- 9
n <- 165

cat(sprintf("Frequentist estimator for p: %.3f", y/n))
```

***(f) find a bayesian estimator for p, assuming both a Beta(1,10) prior for p, and assuming the posterior probability of the older measurement as the prior for the new one.***

Assuming a $Beta(1,10)$ prior:

$$\begin{cases}\alpha=1+9=10\\ \beta=10+165-9=166\end{cases}$$

```{r}
alpha <- 10
beta  <- 166

cat(sprintf("Bayesian estimator for p: %.3f", alpha/(alpha+beta)))
```

Instead, assuming the previous posterior as a prior:

$$\begin{cases}\alpha=12+9=21\\ \beta=115+165-9=271\end{cases}$$

```{r}
alpha <- 21
beta  <- 271

cat(sprintf("Bayesian estimator for p: %.3f", alpha/(alpha+beta)))
```

***(g) find the bayesian estimator for p, the posterior mean and variance, and a 95% credible interval***

```{r}
alpha <- 10
beta  <- 166
var_p <- (alpha*beta) / ((alpha+beta)*(alpha+beta)*(alpha+beta+1))

cat(sprintf("Considering a Beta(1, 10) prior:\n"))
cat(sprintf("Variance: %.5f\n", var_p))
cat(sprintf("95%% credibility interval: [%.3f, %.3f]", qbeta(0.025, alpha, beta), qbeta(0.975, alpha, beta)))



alpha <- 21
beta  <- 271
var_p <- (alpha*beta) / ((alpha+beta)*(alpha+beta)*(alpha+beta+1))

cat(sprintf("\n\nConsidering the previous posterior as prior:\n"))
cat(sprintf("Variance: %.5f\n", var_p))
cat(sprintf("95%% credibility interval: [%.3f, %.3f]", qbeta(0.025, alpha, beta), qbeta(0.975, alpha, beta)))
```

***(h) test the hypothesis***

$$H_0:p=0.1 \ \ \text{versus} \ \ H_1:p\neq0.1$$

***at 5% level of significance with both the frequentist and bayesian approach***

```{r}
binom.test(x = y, n = n, p = 0.1)
```

While $H_0$ is accepted considering the frequentist approach, the bayesian approach depends on the prior chosen: with the previous posterior $0.1$ is inside the $95%$ confidence interval, but with the $Beta(1,10)$ prior that is not the case, so the hypothesis $H_0$ is not accepted.

# Exercise 4

```{r}
# Data
y <- 11
n <- 116
data_list <- list(n = n, y = y)

# Model string for stan
model_string  = 
  " 
    data {
      int <lower=0> n; 
      int <lower=0> y;
    } 
    
    parameters { 
      real <lower=0, upper=1> p; 
    } 
    
    model { 
      p ~ beta (1, 10);
      y ~ binomial (n, p); 
    }
  "
```

```{r}
stan_m  <- stan_model(model_code = model_string)
stanFit <- sampling(object = stan_m,
                    data = data_list, 
                    chains = 2, 
                    iter = 10000, 
                    warmup = 1000, 
                    thin = 1)
```

```{r}
# Plot of the distribution
hist_plot <- mcmc_hist(stanFit, pars = c("p"), binwidth = 0.01) +
              labs(title = "Posterior histogram of p",
                   x = "p",
                   y = "Frequency") +
              theme_minimal() +
              theme(plot.title = element_text(hjust = 0.5))

print(hist_plot)

# Print of the results
summary_df <- as.data.frame(summary(stanFit)$summary)
par_values <- summary_df["p", ]
  
cat(sprintf("\n\n\n"))
cat(sprintf("Mean (also bayesian estimator for p): %.4f\n", par_values["mean"]))
cat(sprintf("Variance: %.4f\n", par_values["sd"]^2))
cat(sprintf("Median: %.4f\n", par_values["50%"]))
cat(sprintf("95%% credibility interval: [%.4f, %.4f]\n", par_values["2.5%"], par_values["97.5%"]))
```
